---
title: "Using Random Forests to Predict Injuries"
output: html_document
runtime: shiny
---
# Exploring Collisions With Random Forests 


## Setup

To start we load the packages we will be using.
Then we load our data from a CSV.
```{r setup, warning=FALSE, message=FALSE}

library("rpart")
library('plotly')
library('shiny')
library('leaflet')
library('stringr')
library('dplyr')
library('randomForest')

col <- read.csv('../data/SDOT_Collisions.csv',stringsAsFactors=FALSE )

```



## Cleaning the Data

Our data set is very large, with over 40 columns. In order to keep ourselves from overfitting we have to pick only a few 
of these variables to actully use in our algorithm.We do that by creating a new dataFrame called data and giving it the variables we want. 


```{r  warning=FALSE, message=FALSE }

data = data.frame(col$INJURIES,col$DISTANCE,col$VEHCOUNT,col$PERSONCOUNT, col$ST_COLCODE, col$WEATHER,stringsAsFactors=FALSE )


```

Now that we have the data we want to use we have to clean it.
In order for our algorithm to process the data, we need to make a few changes to it.

First of all we change all injuries from a numeric scale of how many people were hurt, to a bianary scale of just 1 for there were injuries or 0 there was none.

Then we have to change our string variables to numbers by assinging each weather conditon to its own unique number.

Once that is done have to remove rows with null or missing values.
```{r warning=FALSE, message=FALSE}
data$col.INJURIES <- replace(data$col.INJURIES , data$col.INJURIES != "0", 1)
data$col.INJURIES <- replace(data$col.INJURIES , data$col.INJURIES != "1", 0)

data$col.WEATHER <- replace(data$col.WEATHER ,data$col.WEATHER == "Raining", 0)
data$col.WEATHER <- replace(data$col.WEATHER ,data$col.WEATHER == "Overcast", 1)
data$col.WEATHER <- replace(data$col.WEATHER ,data$col.WEATHER == "Clear or Partly Cloudy", 2)
data$col.WEATHER <- replace(data$col.WEATHER ,data$col.WEATHER == "Unknown", 3)
data$col.WEATHER <- replace(data$col.WEATHER ,data$col.WEATHER == "Other", 4)
data$col.WEATHER <- replace(data$col.WEATHER ,data$col.WEATHER == "Snowing", 5)
data$col.WEATHER <- replace(data$col.WEATHER ,data$col.WEATHER == "Fog/Smog/Smoke", 6)
data$col.WEATHER <- replace(data$col.WEATHER ,data$col.WEATHER == "Sleet/Hail/Freezing Rain", 7)
data$col.WEATHER <- replace(data$col.WEATHER ,data$col.WEATHER == "Severe Crosswind", 8)
data$col.WEATHER <- replace(data$col.WEATHER ,data$col.WEATHER == "Blowing Sand or Dirt or Snow", 9)
data$col.WEATHER <- replace(data$col.WEATHER ,data$col.WEATHER == "", 3)


data$col.WEATHER <- replace(data$col.WEATHER , is.na(data$col.WEATHER) , 0)
data$col.INJURIES <- replace(data$col.INJURIES ,is.na(data$col.INJURIES), 0)
data$col.DISTANCE <- replace(data$col.DISTANCE ,is.na(data$col.DISTANCE), 0)
data$col.VEHCOUNT <- replace(data$col.VEHCOUNT , is.na(data$col.VEHCOUNT), 0)
data$col.PERSONCOUNT <- replace(data$col.PERSONCOUNT , is.na(data$col.PERSONCOUNT) , 0)
data$col.ST_COLCODE <- replace(data$col.ST_COLCODE ,is.na(data$col.ST_COLCODE ), 0)


data$col.WEATHER <- factor(data$col.WEATHER)
data$col.INJURIES <- factor(data$col.INJURIES)
data$col.DISTANCE <- factor(data$col.DISTANCE)
data$col.VEHCOUNT <- factor(data$col.VEHCOUNT)
data$col.PERSONCOUNT <- factor(data$col.PERSONCOUNT)
data$col.ST_COLCODE <- factor(data$col.ST_COLCODE)


data = na.roughfix(data)
#sample <- sample_n(data, 500)

data$col.WEATHER <- as.numeric(data$col.WEATHER)
data$col.INJURIES <- as.numeric(data$col.INJURIES)
data$col.DISTANCE <- as.numeric(data$col.DISTANCE)
data$col.VEHCOUNT <- as.numeric(data$col.VEHCOUNT)
data$col.PERSONCOUNT <- as.numeric(data$col.PERSONCOUNT)
data$col.ST_COLCODE <- as.numeric(data$col.ST_COLCODE)

```

## Splitting data into Train/Test

Now that we have our data cleaned we want to split it into two differenct parts.

we put 70 percent of the data into the training set, and the other 30% we save for testing data.
We do this to keep the testing set away from processing so we can use it for comparisons later. 
```{r cars, warning=FALSE}

dt = sort(sample(nrow(data), nrow(data)*.7))

#sampling data for train/test set
trainindex <- sample(data, trunc(length(data)/2))

#Make training data set
train<-data[dt,]

#Make Testing data
test<-data[-dt,]

```

## Random Forest and classification

We already split the data training and testing so now we have to "Train" our algorithm with our training data. 
To do this we make a Random Forest classifier. In our classifier we fead it our training data and also the specific variables we want our algorithm to look at.(Distance of crash, car count, person count, road condition and weather)

```{r pressure,  warning=FALSE}
#Creates Random FOrest Classifier
class <- randomForest(col.INJURIES ~ col.DISTANCE + col.VEHCOUNT + col.PERSONCOUNT + col.ST_COLCODE + col.WEATHER,
                     data = train, importance = TRUE,  ntree = 100)

```

## Prediction

Now that our algorithm has "Learned" what collisions usuly cause injuries, we will have it predict if their was an injury for each collsion in the TESTING data. We use the prediciton function to do this and we just give it our classifier and the testing data.
Then we compare it to the actual data that shows if there was an injury or not and have it compute the error. 

```{r prediciton, warning=FALSE}
#Uses Classifier to predict wheather somone was hurt in a collision or not
pre <- predict(class,test) 

#Gets values of whether an injury occured or not
actual <- test$col.INJURIES

#Finds Root Mean Squared Error to get effectivness of algorithm 
preAct <- data.frame(pre,actual)
rsq <- 1-sum((actual-pre)^2)/sum((actual-mean(actual))^2)
rmse <- rsq <- sqrt( mean( (actual-pre)^2 , na.rm = TRUE ) )

```

## Conclusion
61.2% of the time our algorithm predicted correctly that there was an injury or not an injruy, based on the data we gave it
regarding each collision. 
